Survival in chessland

### 2019-02-28 trunk/chess/papers/survival.tex
Simultaneously compute stats for multiple criteria, since reading and executing games is basically the entire cost
Some light tweaks to survival paper

\begin{itemize}
\item Black and white are probably not significantly different. That
  is, the a1 and a8 rooks probably have about the same survival
  chances (it's known that white has a slight statistical advantage
  but it is probably only around 1\%). So these guesses will be
  written about white's pieces.
\item The d2 and e2 pawns are very active in common openings, and
  are frequently captured as part of those openings. I think they
  are the least likely to survive overall.
\item Bishops and knights are often involved in the opening and
  midgame, and often exchanged nonchalantly. I think they all have
  relatively low survival chances.
\item Although the queen is very valuable, a queen exchange is often
  forced for games that enter the endgame.
\item Rooks tend to be late-game pieces, because they are difficult
  to get out of their corners (and at most one can be activated
  by the fastest method, castling) and are relatively valuable.
\item This leaves the non-central pawns. These are the hardest to
  predict, and they are hard to think about (at least for me) because
  when e.g. the a2 pawn recaptures the b3 pawn that it supported, I
  just think of this as the b3 pawn. Of these pawns, b2 and g2 are
  somewhat weak because they are undefended once the bishop is
  developed (cf. the famous ``poison pawn'' at b2). On the other hand,
  in the fianchetto configuration, this pawn is very strong and often
  survives the entire game without leaving the third rank. Since pawn
  chains usually progress towards the middle of the board, the a2 pawn
  is more likely to be supporting than supported. This both leaves it
  weak to capture, but prone to recapturing. Outside pawns block one's
  own rook, although for this same reason they often clear the file by
  capturing (and so survive). They are also commonly used to push into
  a well-defended king's territory (e.g. in the fianchetto); kingside
  castling is more common, so this means that the h pawns are often
  lost to this fate.
\end{itemize}

This leaves my final ranking, from most surviving to most dead:

\begin{enumerate}
\item  f2
\item  c2
\item  g2
\item  a2
\item  h2
\item  b2,
\item  h1 rook,
\item  a1 rook,
\item  king,
\item  queen,
\item  f1 bishop,
\item  c1 bishop,
\item  g1 knight,
\item  b1 knight,
\item  e2,
\item  d2,
\end{enumerate}

As already copped to, while the author is an aficionado and also knows
how to spell the difficult word aficionado without spell-check, he is
not good at chess. A few drinking buddies with varying chessperience were
also consulted for their wagers.

\subsection{Ben}\footnote{Ben does not prefer to use the shift key.}

edge pawns almost never played til endgame let alone traded off

ph
pa

not quite sure where these should go (pb more likely to see play in
queenside minority attacks in k-side castle games?)

pf
pg
pb

rook play more likely to be active on q side than on k side (also the
classic Nxc7 fork in low rank play), but overall more likely to stay
tucked away compared to q

Rh
Ra

i think IQP positions are more likely than not saccing e in e4 openings
but on the other hand d is often traded off in e4 openings while vice
versa is not as true

pe
pd

q probably involved in many checkmates (low ranked play) or resignations
before traded off (high ranked play)

Q

just randomly guessing k dies in about 1/3 of games, times 1/2 for 2
sides

K

this pawn is a super goner (sicilian, QGA, ...)

pc

most doomed seem to be the minor pieces as i'd guess at least half of
them get traded off on c/f/3/6 or e/d/4/5 in near every game so

Bc
Bf
Nb
Ng

\subsection{Jim}

Most-to-least-survival hero tier list for chess (patch 1.0):

1.: King
 -- If I estimate that about 2/3rds of all regular pieces are captured in an average game, and the probability of any non-king piece being captured is uniform, then the king is clearly the most likely to survive. (I'm going to break symmetry here and rank black king less likely to survive than white king.)

2. Both Rooks
 -- Kept in reserve for castling purposes.

3. The A, B, G, and H pawns
 -- maybe people will forget to move them because they are far from the center.

4. Both Knights
 -- They are slippery, but they often get deep into enemy territory quickly.

5. The C,D,E,F pawns
 -- Moved forward to release various more important pieces $\Rightarrow$ more likely to die.

6. Both Bishops
 -- https://youtu.be/gDnE-5lD7w8

7. The Queen
 -- A high-value target, seems unlikely to survive.

This list was assembled for chess.report by ix, who only barely
understands the rules of chess and rarely plays. This is probably why
the justifications get increasingly nonsensical.

\subsection{David}

Here's my ranking, along with the probabilities of survival that I guess for each piece:

white g-pawn: 0.72
black g-pawn: 0.72
white b-pawn: 0.69
black a-pawn: 0.66
black b-pawn: 0.65
white king: 0.65
white a-pawn: 0.64
black king: 0.55
black f-pawn: 0.54
white f-pawn: 0.54
black h-pawn: 0.53
white h-pawn: 0.53
white kingside rook: 0.52
black kingside rook: 0.52
white queenside rook: 0.51
black queenside rook: 0.51
white c-pawn: 0.49
black c-pawn: 0.44
white light-squared bishop: 0.33
black light-squared bishop: 0.33
white dark-squared bishop: 0.32
black dark-squared bishop: 0.32
white queen: 0.31
black queen: 0.30
white kingside knight: 0.29
black kingside knight: 0.29
white queenside knight: 0.28
black queenside knight: 0.28
white e-pawn: 0.19
black d-pawn: 0.18
white d-pawn: 0.17
black e-pawn: 0.16

\subsection{William}

Okay, this is pretty much off the cuff, but let's say, just for white,
from most to least likely to survive:

Queenside Rook
B Pawn
A Pawn
Kingside Bishop
C Pawn
Queen
King
H Pawn
Queenside Bishop
F Pawn
G Pawn
Queenside Knight
Kingside Rook
Kingside Knight
D Pawn
E Pawn

Some motivated reasoning:

I figure the king has got to be somewhere near the middle of the pack
since he dies in half of games featuring a winner---but with slightly
higher-than-even odds of surviving, since some games end in a draw.
I’m probably mixing up means and medians here somehow..

...
King
...

I’m gonna assume castling happens more often on the King’s side, so
let’s give Kingside Rook and F, G, and H Pawns a better shot than
their fellows on the left. But maybe it should actually be worse,
since if they die, it's because they failed to protect the king. Plus,
having heard the tip about C Pawn loud and clear, I’m gonna assume
that bad boy most often becomes a new Queen, which means he gets more
survival points than the real Queen herself.

...
C Pawn
Queen
A Pawn
B Pawn
Queenside Rook
King
F Pawn
G Pawn
H Pawn
Kingside Rook
...

...
D Pawn
E Pawn

Randomizing within these constraints gives us our starting point. Then
the wildcard Bishops and Knights get randomly distributed through what
remains to come up with this final answer shown above.

@@

\begin{figure}[htb]
  \begin{center}
    % To generate this pdf, I ran maketable to make
    % piece-survival.svg, then printed to PDF using Chrome, then
    % cropped it in Illustrator. Illustrator can open the SVG but
    % doesn't have the unicode chess pieces in any font I could find.
    \includegraphics[width=0.95 \linewidth]{piece-survival-export}
  \end{center}\vspace{-0.1in}
  \caption{Survival probabilities for each of the 32 pieces in standard
    chess, in 500 million games. The number is the mean survival rate
    across all samples. The vertical position of the dot is this mean
    rate, with a line drawing the span between the smallest and largest
    sample bucket (this is usually a very tiny range).
    Horizontal position is purely presentational, to avoid overlap.}
  \label{fig:piece-survival}
\end{figure}

The basic survival chances appear in Figure~\ref{fig:piece-survival}.
Indeed, many pieces are more likely to survive than the kings. Even as
black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
survival rate. Across the board, the survival chances for a white
piece and its black twin are similar, usually with a small edge to
white. Notable exceptions are the \knight g (the overall most doomed
piece), and both white bishops, which die more than their
Schwarzdoppelgangers. The \pawn e is vastly more dead than \Pawn e.
Note somewhat satisfyingly that the \pawn c has the highest variance;
this was the most controversial among the drinking buddies
(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed
pawn in the popular Queen's Gambit (1. d4 d5 2. c4), where accepting
and declining the pawn are both popular and sound responses. This may
be a good example of a piece that has substantially different survival
rates in different opening preferences. Since the variance is
otherwise extremely low, I only report means for the remainder of the
paper.

Despite my impression that many games end in a draw, ties are actually
rare in the lichess database. In January 2018, only 3.8\% of games
were drawn;
% (8922275 white, 8329221 black, 682042 draw)
as a result, the survival chances for the kings are both close to
50\%. Although the database contains games in many time formats and
with all varieties of human skill (including over a thousand games by
Magnus Carlsen, the world champion and highest rated player of all
time\footnote{Although to be fair, his username ``DrDrunkenstein''
  suggests he may not play at full strength.}), blitz ($\sim 5$ minutes
per side) and bullet ($\sim 1$ minute) games are predominant. In
Section~\ref{sec:comparative} I show some slices of the data for
comparison. % XXX could summarize whether this is a real effect or
% not, once I have the slices!

@@

Other than the bishops---which have no legal way to reach half of the
squares---every piece ends on every square in at least a thousand
games. So we have enough samples to have reasonable confidence in our
statistics, even for the most unlikely odysseys. The least mobile
pieces are the pawns, who can technically reach any square by
promoting, but are usually confined to cones emanating from their
start squares. The overall rarest fate is for \pawn f2 to die on the
a7 square, which only happened 1,244 times (however, it survived on
this square 31,438 times). This square is actually reachable without
promoting, but it would need to capture 5 times in order to get
there, which seems quite unlikely! There may even be a hidden
achievement for reaching this square this way!
% TODO: Perhaps could find a game where this happened?!
Aside from pawns, the weirdest fate is for \knight g to die on h1,
which happened 47,307 times. Corners are of course fairly garbage for
the knight, although it is twice as likely to survive on this square.

% There may even be hidden achievements for reaching these squares!

That said, there are characteristic patterns for each of the pieces,
which make sense given their starting positions and patterns of
movement. You could probably guess the piece just by looking at one of
the heat maps below, although---spoiler alert---the piece is just
listed right there and they are in order. Two independent things are
communicated in these graphics: The chance that the piece ends the
game on some square (alive or dead), and its survival chances there.
In each map, a darker background color indicates that the piece ends
on that square more often. The shade is based on the rank (64/64 black
is most common, 63/64 black is next most common, etc.) rather than the
absolute probability, since otherwise the graphic looks boring. The
number on the square is the percentage survival of the piece when it
is last seen (either being captured or surviving to the end of the
game) on that square. The four squares with the highest survival rates
are underlined for your convenience.

@@

\section{Guesses and slices}
%  
%  \begin{tabular}{|l|l|}
%  \hline
%    tom & actual \\
%   \hline
%   \Pawn f    &  \\
%   \Pawn c    &  \\
%   \Pawn g    &  \\
%   \Pawn a    &  \\
%   \Pawn h    &  \\
%   \Pawn b    &  \\
%   \Rook h    &  \\
%   \Rook a    &  \\
%   \King      &  \\
%   \Queen     &  \\
%   \Bishop f  &  \\
%   \Bishop c  &  \\
%   \Knight g  &  \\
%   \Knight b  &  \\
%   \Pawn e    &  \\
%   \Pawn d    &  \\
%  \hline
%  \end{tabular}
%

>>>>

\begin{itemize}
\item Black and white are probably not significantly different. That
  is, the a1 and a8 rooks probably have about the same survival
  chances (it's known that white has a slight statistical
  advantage~\cite{chesstego} but it is probably only around 1\%). So
  these guesses will be written about white's pieces.
\item The d2 and e2 pawns are very active in common openings, and
  are frequently captured as part of those openings. I think they
  are the least likely to survive overall.
\item Bishops and knights are often involved in the opening and
  midgame, and often exchanged nonchalantly. I think they all have
  relatively low survival chances.
\item Although the queen is very valuable, a queen exchange is often
  forced for games that enter the endgame.
\item Rooks tend to be late-game pieces, because they are difficult
  to get out of their corners (and at most one can be activated
  by the fastest method, castling) and are relatively valuable.
\item This leaves the non-central pawns. These are the hardest to
  predict, and they are hard to think about (at least for me) because
  when e.g. the a2 pawn recaptures the b3 pawn that it supported, I
  just think of this as the b3 pawn. Of these pawns, b2 and g2 are
  somewhat weak because they are undefended once the bishop is
  developed (cf. the famous ``poison pawn'' at b2). On the other hand,
  in the fianchetto configuration, this pawn is very strong and often
  survives the entire game without leaving the third rank. Since pawn
  chains usually progress towards the middle of the board, the a2 pawn
  is more likely to be supporting than supported. This both leaves it
  weak to capture, but prone to recapturing. Outside pawns block one's
  own rook, although for this same reason they often clear the file by
  capturing (and so survive). They are also commonly used to push into
  a well-defended king's territory (e.g. in the fianchetto); kingside
  castling is more common, so this means that the h pawns are often
  lost to this fate.
\end{itemize}

The final ranking that I predict, from most surviving to most dead:
\pawn f, \pawn c, \pawn g, \pawn a, \pawn h, \pawn b, \rook h, \rook a,
\king, \queen, \bishop f, \knight g, \pawn e, \pawn d.

As already copped to, while the author is an aficionado and also knows
how to spell the difficult word aficionado without spell-check, he is
not good at chess. A few drinking buddies with varying chessperience
were also consulted for their wagers; these are compared to each other
and to the actual results in Section~\ref{sec:guesses}.

@@

\begin{figure}[t]
  \begin{center}
    % To generate this pdf, I ran maketable to make
    % piece-survival.svg, then printed to PDF using Chrome, then
    % cropped it in Illustrator. Illustrator can open the SVG but
    % doesn't have the unicode chess pieces in any font I could find.
    \includegraphics[width=0.95 \linewidth]{piece-survival-export}
  \end{center}\vspace{-0.1in}
  \caption{Survival probabilities for each of the 32 pieces in standard
    chess, in 500 million games. The number is the mean survival rate
    across all samples. The vertical position of the dot is this mean
    rate, with a line drawing the span between the smallest and largest
    sample bucket (this is usually a very tiny range).
    Horizontal position is purely presentational, to avoid overlap.}
  \label{fig:piece-survival}
\end{figure}

The basic survival chances appear in Figure~\ref{fig:piece-survival}.
Indeed, many pieces are more likely to survive than the kings. Even as
black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
survival rate. Across the board, the survival chances for a white
piece and its black twin are similar, usually with a small edge to
white. Notable exceptions are the \knight g (the overall most doomed
piece), and both white bishops, which die more than their
Schwarzdoppelgangers. The \pawn e is vastly more dead than \Pawn e.
Note somewhat satisfyingly that the \pawn c has the highest variance;
this was the most controversial among the drinking buddies
(Section~\ref{sec:guesses}). Note that \pawn c is the sacrificed
pawn in the popular Queen's Gambit (1. d4 d5 2. c4), where accepting
and declining the pawn are both common and sound responses. This may
be a good example of a piece that has substantially different survival
rates in different opening preferences. Since the variance is
otherwise extremely low, I only report means for the remainder of the
paper.

Despite my impression that many games end in a draw, ties are actually
rare in the lichess database. In January 2018, only 3.8\% of games
were drawn;
% (8922275 white, 8329221 black, 682042 draw)
as a result, the survival chances for the kings are both close to
50\%. Although the database contains games in many time formats and
with all varieties of human skill (including over a thousand games by
Magnus Carlsen, the world champion and highest rated player of all
time\footnote{Although to be fair, his username ``DrDrunkenstein''
  suggests he may not play at full strength.}), blitz ($\sim\! 5$
minutes per side) and bullet ($\sim\! 1$ minute) games are
predominant. Nonetheless, the results are fairly robust across
different time formats and skill levels. In Section~\ref{sec:slices} I
show some slices of the data for comparison.

@@

Other than the bishops---which have no legal way to reach half of the
squares---every piece ends on every square in at least a thousand
games. So we have enough samples to have reasonable confidence in our
statistics, even for the most unlikely odysseys. The least mobile
pieces are the pawns, who can technically reach any square by
promoting, but are usually confined to cones emanating from their
start squares. The overall rarest fate is for \pawn f2 to die on the
a7 square, which only happened 1,244 times (however, it survived on
this square 31,438 times). This square is actually reachable without
promoting, but it would need to capture 5 times in order to get
there, which seems quite unlikely! There may even be a hidden
achievement for reaching this square this way!
% 12 times for titled players (in new dataset too), although there are
% a few things that are even rarer, like dying on a8.
% TODO: Perhaps could find a game where this happened?!
Aside from pawns, the weirdest fate is for \knight g to die on h1,
which happened 47,307 times. Corners are of course garbage for the
knight, although it is twice as likely to survive on this square.

There are characteristic patterns for each of the pieces, which make
sense given their starting positions movement rules. You could
probably guess the piece just by looking at one of the heat maps
below, although---spoiler alert---the piece is just listed right there
and they are in order. Two independent things are communicated in
these graphics: The chance that the piece ends the game on some square
(alive or dead), and its survival chances there. In each map, a darker
background color indicates that the piece ends on that square more
often. The shade is based on the rank (64/64 black is most common,
63/64 black is next most common, etc.) rather than the absolute
probability, since otherwise the graphic looks boring. The number on
the square is the percentage survival of the piece when it is last
seen (either being captured or surviving to the end of the game) on
that square. The four squares with the highest survival rates are
underlined for your convenience.

\clearpage

@@

\clearpage

Aside from admiring the groovy pictures, the data can also be used as
the basis of an algorithm for playing chess. There are several things
to try; we could move pieces towards squares where they are more
likely to survive than die (keeping pieces alive is good), or just
towards squares where they are more likely to be positioned at the end
of the game (our moves are more likely to resemble real chess moves
because they put pieces in their proper places). Alas, it turns out
that these are bad approaches to chess, both because they are boring
(most pieces are actually most comfortable in their starting
positions) and because they perform badly against even weak
opponents~\cite{eloworld}.

\section{Guesses and slices} \label{sec:guesses}

Like all good scientific research, I explicitly compare the actual
results to the hypotheses gathered before the experiment; this is a
hygenic and humbling exercise. Figure~\ref{fig:buddies} compares the
ranking across all games (slice {\bf Actual}) to the author author
(slice {\bf Tom}) and his drinking buddies (slices {\bf Ben}, {\bf
  Jim}, {\bf David}, {\bf William}). There are a number of different
reasonable ways to measure the accuracy of this type of position; a
very simple one is the sum of the absolute differences in rank for
each piece (e.g. if in one ranking \king is \#3, and in the other \#5,
then this contributes 2 to the total error). By this metric, Ben has
the best prediction (98 error), followed by David (138) and Tom (148).
Tom and David had the most similar predictions (116) and Jim and
William the most different (312). The expected error between two
completely random permutations is about 341, so all of these guesses
are significantly better than chance. Note in the actual ranking, many
pieces have very similar survival probabilities, and many guesses are
ambivalent about groups of pieces. Weighting each rank difference
equally is therefore an oversimplification. It would have been better
to ask each participant to give probabilities, as David did; this
would give us more sensitive error metrics and more opportunities to
spend the afternoon making visualizations.

Several drinking buddies gave rationales for their hypotheses (mine
appear in Section~\ref{sec:hypotheses}).

\medskip
{\bf Ben} does not prefer to use the shift key, a typographic quirk I
replicated faithfully here even though it burns my eyes:

\begin{quote}
edge pawns almost never played til endgame let alone traded off (\pawn
h, \pawn a)

not quite sure where these should go (pb more likely to see play in
queenside minority attacks in k-side castle games?) (\pawn f, \pawn g, \pawn b)

rook play more likely to be active on q side than on k side (also the
classic Nxc7 fork in low rank play), but overall more likely to stay
tucked away compared to q (\rook h, \rook a)

i think IQP positions are more likely than not saccing e in e4 openings
but on the other hand d is often traded off in e4 openings while vice
versa is not as true (\pawn e, \pawn d)

q probably involved in many checkmates (low ranked play) or resignations
before traded off (high ranked play) (\queen)

just randomly guessing k dies in about 1/3 of games, times 1/2 for 2
sides (\king)

this pawn is a super goner (sicilian, QGA, ...) (\pawn c)

most doomed seem to be the minor pieces as i'd guess at least half of
them get traded off on c/f/3/6 or e/d/4/5 in near every game so (\bishop c, \bishop f, \knight b, \knight g)
\end{quote}

\begin{figure}[t]
  \includegraphics[width=\linewidth]{hypotheses}
  \caption{
    Piece rankings (from most surviving to most dead); either a
    human or hypothesis or the actual results across all games. 
    The {\bf actual} column appears multiple times so that each human
    gets a chance to be adjacent to it; this makes for the easiest
    visual comparison. Lines connect the same piece in adjacent columns,
    and are darker if the pairs have more different ranks.
  } \label{fig:buddies}
\end{figure}

{\bf Jim} ``barely understands the rules of chess'' and ``rarely plays.'' His justifications get ``increasingly nonsensical:''

\begin{quote}
Most-to-least-survival hero tier list for chess (patch 1.0):

1.: King --- If I estimate that about 2/3rds of all regular pieces are
captured in an average game, and the probability of any non-king piece
being captured is uniform, then the king is clearly the most likely to
survive. (I'm going to break symmetry here and rank black king less
likely to survive than white king.)

2. Both Rooks --- Kept in reserve for castling purposes.

3. The A, B, G, and H pawns --- maybe people will forget to move them because they are far from the center.

4. Both Knights --- They are slippery, but they often get deep into enemy territory quickly.

5. The C,D,E,F pawns --- Moved forward to release various more important pieces $\Rightarrow$ more likely to die.

6. Both Bishops --- https://youtu.be/gDnE-5lD7w8

7. The Queen --- A high-value target, seems unlikely to survive.
\end{quote}

{\bf David} simply provided a ranking, along with survival
probabilities, in typical understated style.

\medskip
{\bf William} notes that his guesses are
``pretty much off the cuff,'' but provides some motivated reasoning:

\begin{quote}
I figure the king has got to be somewhere near the middle of the pack
since he dies in half of games featuring a winner---but with slightly
higher-than-even odds of surviving, since some games end in a draw.
I'm probably mixing up means and medians here somehow..

I'm gonna assume castling happens more often on the King's side, so
let's give Kingside Rook and F, G, and H Pawns a better shot than
their fellows on the left. But maybe it should actually be worse,
since if they die, it's because they failed to protect the king. Plus,
having heard the tip about C Pawn\footnote{I believe the ``tip'' here
  was that I described the rest of us (William was the last
  respondant) as disagreeing most on \pawn c. I think William
  misinterpreted this ``tip.''} loud and clear, I'm gonna assume that
bad boy most often becomes a new Queen, which means he gets more
survival points than the real Queen herself.

Randomizing within these constraints gives us our starting point. Then
the wildcard Bishops and Knights get randomly distributed through what
remains to come up with this final answer shown above.
\end{quote}

\subsection{Slices} \label{sec:slices}

\begin{figure}
  \includegraphics[width=\linewidth]{slices}
  \caption{
    Ranks and survival probabilities for different subsets of games.
    The same piece in adjacent rows is connected to highlight
    differences in the ranking, as before. The time formats all
    exhibit similar ranking with only small perturbations. Games
    including a titled player (rightmost column) are the most
    different, although we have far fewer samples in this set,
    so variance becomes significant.
  } \label{fig:slices}
\end{figure}

The survival probabilities differ depending on the conditions of the
game; Figure~\ref{fig:slices} compares some of those slices. Here the
{\bf All} slice is the same as the {\bf Actual} column in
Figure~\ref{fig:buddies}, and consists of all acceptable games in the
database.

The {\bf Titled} slice includes only games where at least one of the
players has an official title (Grandmaster, International Master, FIDE
Master, etc.\footnote{Lichess used to award the LM ``Lichess Master''
  title to notable players on the site; this title is excluded from
  the sample.}). These games have high-quality play, but far fewer
samples (``only'' 3.4 million). This set exhibits significant
variance; for example, the survival rate of \pawn c ranges from
$38$--$46\%$. This is both because of the small sample size and the
bucketization by player name; there are few enough titled players that
an individual's preference in openings and style of play changes the
values of their entire bucket. I caution against reading too much into
this column. It seems we can at least conclude that these games tend
to be much bloodier (Kings aside, survival rates are lower across the
board); top players are less likely to fall for traps early in the
game, perhaps more willing to sacrifice material, and more likely to
play into endgames where almost every piece is exchanged. If being a
chesspiece to the death, you do {\em not} want to have a Grandmaster
playing the game!

% XXX give % error % bounds?

On the other hand, the other slices all have enough samples that the
variance is minimal. These slices, {\bf Bullet} (151,261,707 games),
{\bf Blitz} (236,050,938 games), {\bf Rapid} (98,606,558) and {\bf
  Classical} (13,886,352) are different time control formats. Games on
Lichess are played with a starting clock (per side) and an increment
added to the clock after each player's move. The game is classified
according to the estimated total time: The starting time plus $40 \times$
the increment (with the idea that an average game has 40 moves per
side); this is the same formula that Lichess uses. A bullet game is
when the total time per side is between 30~seconds and 3~minutes;
blitz is between 3 and 8~minutes; rapid is between 8 and 25~minutes;
classical is any more than this (including untimed games). Each of
these slices has enough samples that the variance is very low.
Here we see that the ranking is rather stable across the range of
time formats, which was not what I expected. This should increase
our confidence that the results are really inherent to chess, not
just the particulars of this data set.
